---
title: "Prerequisits"
author: ''
date: ''
output:
  rmarkdown::html_document:
    highlight: default
    theme: journal
  rmarkdown::pdf_document: null
subtitle: Manual for R package `reservoir`
---


## Step 1: Get R and R Studio

***R*** is an open source, statistical programming environment, popular with academics in a variety of fields, including hydrology and water resources engineering.

You can download and install the latest version of R from https://cran.r-project.org/.

***R Studio*** is a powerful Integrated Development Environment (IDE) for R. Whislt an IDE is not absolutely necessary to use `reservoir`---or indeed any R library---it makes life lot easier. There are various IDEs for R, but R Studio is the most popular and its capabilities extend well beyond writing basic R code (this manual was written using R Studio, for instance).

Download and install the latest version of R-Studio from https://www.rstudio.com/products/rstudio/download/.

Some basic knowledge of R is recommended (though not absolutely necessary) for following this manual and working with *reservoir*. Here are some popular introductory courses for first-time users:

* [DataCamp's free introduction to R](https://www.datacamp.com/courses/free-introduction-to-r)

* [ComputerWorld's Beginners Guide to R](http://www.computerworld.com/article/2497143/business-intelligence/business-intelligence-beginner-s-guide-to-r-introduction.html)

* [swirl](http://swirlstats.com/)


## Step 2: Get the *reservoir* package

You can download reservoir directly from the R console. Two versions will be available at any time: a stable release version (recommended), stored on the Comprehensive R Archive Network (CRAN), and a development version, stored on Github.

You can get the release (CRAN) version using `install.packages`:

```{r, eval=FALSE}
install.packages("reservoir")
```

New developments to *reservoir* won't be always be available on CRAN. If you want to make sure you have all new features, you can install the development version using `devtools`:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("swd-turner/reservoir")
```

Once installed, load *reservoir* into the working environment using:
```{r, eval=FALSE}
library("reservoir")
```

Access package documentation using:
```{r, eval=FALSE}
?reservoir
```




